Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-03-27 _tex/paper2.tex
Built site for gh-pages

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  review,
  1p]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Psychometrika}

\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation},
  pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
  pdfkeywords={causal inference, directed acyclic graphs, structural
causal models, bayesian statistical methods, thurstonian
model, comparative judgement, probability, statistical modeling},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{abstract}
This study revisits Thurstone's law of comparative judgment (CJ),
focusing on two prominent issues of traditional approaches. First, it
critiques the heavy reliance on Thurstone's Case V assumptions and, by
extension, the Bradley-Terry-Luce (BTL) model when analyzing CJ data.
Specifically, the study raises concerns about the assumptions of equal
discriminal dispersions and zero correlation between the stimuli. While
these assumptions simplify the trait measurement process, they may fail
to capture the complexity of CJ data, potentially leading to unreliable
and inaccurate trait estimates. Second, the study highlights the
apparent disconnect between CJ's trait measurement and hypothesis
testing processes. Although separating these processes simplifies the
analysis of CJ data, it may also undermine the reliability of various
statistical results derived from the measurement.

@@

Finally, the study emphasizes the relevance of this extension for
contemporary empirical CJ research, particularly stressing the need for
bespoke CJ models tailored to the specific assumptions of the
experiments and data. It also lays the foundation for broader
applications, encouraging researchers across the social sciences to
adopt more robust and interpretable methodologies.
\end{abstract}

@@

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues of traditional approaches that can undermine the
reliability and validity of CJ's trait estimates
\citep[pp.~2]{Perron_et_al_2015}. First, it critiques the heavy reliance
on Thurstone's Case V assumptions \citep{Thurstone_1927b} and, by
extension, the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} when analyzing CJ data.
Specifically, the study raises concerns about the assumptions of equal
discriminal dispersions and zero correlation between the stimuli. While
these assumptions simplify statistical modeling, they may fail to
capture the complexity of CJ data, potentially leading to unreliable and
inaccurate trait estimates. Second, the study highlights the disconnect
between CJ's trait measurement and hypothesis testing processes.
Although separating these processes simplifies the analysis of CJ data,
it may also undermine the reliability of various statistical results
derived from the measurement.

@@

To address this issue and facilitate the practical implementation of the
theory, Thurstone developed five cases derived from this general form,
each progressively incorporating additional simplifying assumptions into
the model. In Case I, Thurstone postulated that pairs of stimuli would
maintain a constant correlation across all comparisons. In Case II, he
allowed multiple judges to undertake comparisons instead of confining
evaluations to a single judge. In Case III, he posited that there was no
correlation between stimuli. In Case IV, he assumed that the stimuli
exhibited similar dispersions. Finally, in Case V, he replaced this
assumption with the condition that stimuli had equal discriminal
dispersions. Table~\ref{tbl-thurstone_cases} summarizes the assumptions
of the general form and the five cases. For a detailed discussion of
these cases and their progression, refer to \citet{Thurstone_1927b} and
\citet[pp.~248--253]{Bramley_2008}.

@@

Notably, while the assumption of equal dispersions simplifies the trait
measurement process, evidence from the CJ literature suggests that the
BTL model may fail to capture the complexity of modern CJ data.
Specifically, the assumption of equal dispersions may not hold when
researchers assess complex traits or heterogeneous stimuli
\citetext{\citealp[pp.~376]{Thurstone_1927a}; \citealp[pp.~273]{Bramley_2008}; \citealp[pp.~678]{Kelly_et_al_2022}},
as these traits and stimuli can introduce judgment discrepancies due to
their unique characteristics
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, the CJ literature may already provide evidence of such
discrepancies, particularly in the form of ``misfit'' statistics. For
instance, \emph{misfit texts} are those whose comparisons result in more
judgment discrepancies than those involving other texts
\citetext{\citealp[pp.~11]{Pollitt_2004}; \citealp[pp.~290]{Pollitt_2012b}}.

These discrepancies may arise from two possible causes. First, the
discriminal processes of misfit texts may substantially overlap with
those of non-misfit texts. Second, misfit texts may represent outlying
observations with characteristics that make them difficult to compare
and position on the trait scale. Regarding the first cause, the overlap
can occur when misfit texts have equal dispersion and share the same
trait space as non-misfit texts or when they exhibit larger discriminal
dispersions. Although the BTL model can distinguish the first scenario,
it is inherently agnostic to the second.
Figure~\ref{fig-discriminal_process} illustrates how an increased
discriminal dispersion can widen the overlap between two texts, leading
to more judgment discrepancies. In the figure, text B has a broader
discriminal dispersion than text A, expanding the overlap between their
discriminal processes, a space where judges struggle to determine which
text possesses more of the trait.

When the assumption of equal dispersions between stimuli does not hold,
significant statistical and measurement issues can arise Specifically,
the BTL model may overestimate the trait's reliability, that is, the
degree to which the outcome accurately reflects the \emph{true}
discriminal differences between stimuli. This overestimation, in turn,
results in spurious conclusions about these differences
\citep[pp.~370]{McElreath_2020} and, by extension, about the underlying
discriminal processes of stimuli. Figure~\ref{fig-dispersion} also
illustrates this scenario when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the \emph{true} discriminal
difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\).

@@

Thus, erroneously assuming zero correlation between stimuli can also
lead to significant statistical and measurement issues. Specifically,
neglecting judges' biases or relevant hierarchical structures can create
dimensional mismatches in the model, leading to the over- or
underestimation of trait reliability \citep[pp.~341,
482]{Ackerman_1989, Hoyle_et_al_2023}. These inaccuracies can result in
spurious conclusions about the discriminal differences
\citep[pp.~370]{McElreath_2020} and, by extension, the underlying
discriminal processes of the stimuli. This issue is illustrated in
Figure~\ref{fig-correlation} when the discriminal difference
distribution of the BTL model follows the thick continuous line
\((\rho = 0)\), while the \emph{true} discriminal difference follows any
discontinuous line where \((\rho \neq 0)\).

Finally, as discussed in the previous section, removing judges based on
misfit statistics risks discarding valuable information and even
introduce bias into the trait estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The direction and
magnitude of these biases remain unpredictable because they depend on
which judges researchers exclude from the analysis. \emph{Misfit judges}
are those whose evaluations deviate substantially from the shared
consensus due to the unique characteristics of either the stimuli or the
judges themselves
\citetext{\citealp[pp.~164-165]{Pollitt_2012a}; \citealp[pp.~289-290]{Pollitt_2012b}; \citealp[pp.~4]{vanDaal_et_al_2016}; \citealp[pp.~20]{Goossens_et_al_2018}}.

@@

Nevertheless, while separating the trait measurement and hypothesis
testing processes simplifies the analysis of CJ data, the statistical
literature cautions against relying solely on the point estimates of BTL
scores to conduct further analyses or hypothesis tests, as this practice
can undermine the resulting statistical conclusions. A key consideration
is that BTL scores are parameter estimates that inherently carry
uncertainty (measurement error). Ignoring this uncertainty can bias the
analysis and reduce the precision of hypothesis tests. The direction and
magnitude of such biases are often unpredictable. Results may be
attenuated, exaggerated, or remain unaffected depending on the degree of
uncertainty in the scores and the actual effects being tested
\citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
Furthermore, the reduced precision in hypothesis tests diminishes their
statistical power, increasing the likelihood of committing type-I or
type-II errors \citep{McElreath_2020}.
Figure~\ref{fig-measurement_error} illustrates these issues,
demonstrating how neglecting measurement error by relying only on
outcome averages can reduce the precision of a predictor's estimated
effect.

@@

\caption{\label{fig-measurement_error}The effect of outcome uncertainty
on the estimation of an effect \(\beta\) linked to the predictor.
Uncertainty increases from left to right.}

@@

Luckily, SCMs and DAGs support identification analysis through two key
advantages. First, regardless of complexity, they can represent various
causal structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components, facilitating
their analysis \citep{McElreath_2020}. Second, they depict causal
relationships in a non-parametric way. This flexibility enables feasible
identification strategies without requiring specification of the types
of variables, the functional forms relating them, or the parameters of
those functional forms \citep[pp.~35]{Pearl_et_al_2016}.

@@

Before incorporating the theoretical principles into the
\emph{conceptual-population model}, it is essential to further define
SCMs. SCMs are formal mathematical models characterized by a set of
\emph{endogenous} variables \(V\), a set of \emph{exogenous} variables
\(E\), and a set of functions \(F\)
\citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous variables are those
whose causal mechanisms a researcher chooses to model \citep{Neal_2020}.
In contrast, exogenous variables represent \emph{errors} or
\emph{disturbances} arising from omitted factors that the investigator
chooses not to model explicitly \citep[pp.~27,68]{Pearl_2009}. Lastly,
the functions, referred to as \emph{structural equations}, express the
endogenous variables as non-parametric functions of other endogenous and
exogenous variables. These functions use the symbol `\(:=\)' to denote
the asymmetrical causal dependence between variables and the symbol
`\(\:\bot\:\)' to represent \emph{d-separation}, a concept akin to
(conditional) independence.

SCM~\ref{fig-cj03_scm} illustrates the first theoretical principles of
the conceptual-population model designed to examine the impact of
different teaching methods on students' writing ability. This SCM
outlines the relationship between the conceptual-population outcome
\((O^{cp}_{hiabjk})\) and several related variables. The subscripts
\(h\) and \(i\) label the students who authored the texts (individuals),
while the indices \(a\) and \(b\) represent the compared texts
(stimuli). The index \(j\) refers to the judge, and \(k\) represents the
judgment index, accounting for experimental conditions where judges
perform the same judgment multiple times, i.e., a \emph{repeated
measures design} \citep[pp.~366-376]{Lawson_2015}. Notably, the indexing
system facilitates comparisons between texts written by the same student
(\(h = i\), \(a \neq b\)) and between texts written by distinct students
(\(h \neq i\), where \(a = b\) is possible). However, it excludes cases
where judges compare a student's text to itself (\(h = i\), \(a = b\)),
as such comparison lacks practical relevance within the CJ framework.

@@

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a \emph{graph}
consisting of nodes connected by edges, where nodes represent random
variables. The term \emph{directed} indicates that edges or arrows
extend from one node to another, indicating the direction of causal
influence. The absence of an edge implies no direct relationship between
the nodes. The term \emph{acyclic} means that the causal influences do
not form loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs conventionally depict observed variables as
solid black circles and unobserved (latent) variables as open circles
\citep{Morgan_et_al_2014}. They also display conditioned variables by
surrounding them with a square. Here, \emph{conditioning} refers to
restricting the statistical analysis to a subset of the population
defined by the values of the given variable \citep[pp.~32]{Neal_2020}.
Although DAGs often omit exogenous variables for simplicity, this
section includes exogenous variables to improve clarity and reveal
potential issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}.

>>>>

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  review,
  1p]{elsarticle}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Psychometrika}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation},
  pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
  pdfkeywords={causal inference, directed acyclic graphs, structural
causal models, bayesian statistical methods, thurstonian
model, comparative judgement, probability, statistical modeling},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{abstract}
This study revisits Thurstone's law of comparative judgment (CJ),
focusing on two prominent issues of traditional approaches. First, it
critiques the heavy reliance on Thurstone's Case V assumptions and, by
extension, the Bradley-Terry-Luce (BTL) model when analyzing CJ data.
Specifically, the study raises concerns about the assumptions of equal
discriminal dispersions and zero correlation between the stimuli. While
these assumptions simplify the trait measurement model, they may fail to
capture the complexity of CJ data, potentially leading to unreliable and
inaccurate trait estimates. Second, the study highlights the apparent
disconnect between CJ's trait measurement and hypothesis testing
processes. Although separating these processes simplifies the analysis
of CJ data, it may also undermine the reliability of various statistical
results derived from these processes.

@@

Finally, the study emphasizes the relevance of this extension for
contemporary empirical CJ research, particularly stressing the need for
bespoke CJ models tailored to the experiments and data assumptions. It
also lays the foundation for broader applications, encouraging
researchers across the social sciences to adopt more robust and
interpretable methodologies.
\end{abstract}

@@

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues of traditional approaches that can undermine the
reliability and validity of CJ's trait estimates
\citep[pp.~2]{Perron_et_al_2015}. First, it critiques the heavy reliance
on Thurstone's Case V assumptions \citep{Thurstone_1927b} and, by
extension, the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} when analyzing CJ data.
Specifically, the study raises concerns about the assumptions of equal
discriminal dispersions and zero correlation between the stimuli. While
these assumptions simplify the trait measurement model, they may fail to
capture the complexity of CJ data, potentially leading to unreliable and
inaccurate trait estimates. Second, the study highlights the disconnect
between CJ's trait measurement and hypothesis testing processes.
Although separating these processes simplifies the analysis of CJ data,
it may also undermine the reliability of various statistical results
derived from these processes.

@@

To address this issue and facilitate the practical implementation of the
theory, Thurstone developed five cases derived from this general form,
each progressively incorporating additional simplifying assumptions. In
Case I, Thurstone postulated that pairs of stimuli would maintain a
constant correlation across all comparisons. In Case II, he allowed
multiple judges to undertake comparisons instead of confining
evaluations to a single judge. In Case III, he posited that there was no
correlation between stimuli. In Case IV, he assumed that the stimuli
exhibited similar dispersions. Finally, in Case V, he replaced this
assumption with the condition that stimuli had equal discriminal
dispersions. Table~\ref{tbl-thurstone_cases} summarizes the assumptions
of the general form and the five cases. For a detailed discussion of
these cases and their progression, refer to \citet{Thurstone_1927b} and
\citet[pp.~248--253]{Bramley_2008}.

@@

Notably, while the assumption of equal dispersions simplifies the trait
measurement model, evidence from the CJ literature suggests that the BTL
model may fail to capture the complexity of modern CJ data.
Specifically, the assumption of equal dispersions may not hold when
researchers assess complex traits or heterogeneous stimuli
\citetext{\citealp[pp.~376]{Thurstone_1927a}; \citealp[pp.~273]{Bramley_2008}; \citealp[pp.~678]{Kelly_et_al_2022}},
as these traits and stimuli can introduce judgment discrepancies due to
their unique characteristics
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, the CJ literature may already provide evidence of such
discrepancies, particularly in the form of ``misfit'' statistics. For
instance, \emph{misfit texts} are those whose comparisons result in more
judgment discrepancies than those involving other texts
\citetext{\citealp[pp.~11]{Pollitt_2004}; \citealp[pp.~290]{Pollitt_2012b}}.

These discrepancies may arise from two possible causes. First, the
discriminal processes of misfit texts may substantially overlap with
those of non-misfit texts. Second, misfit texts may represent outlying
observations with characteristics that make them difficult to compare
and position on the trait scale. Regarding the first cause, the overlap
can occur when misfit texts have equal dispersion and share the same
trait space as non-misfit texts or when they exhibit larger discriminal
dispersions. Although the BTL model can distinguish the first scenario,
it is inherently agnostic to the second.
Figure~\ref{fig-discriminal_process} illustrates how an increased
discriminal dispersion can widen the overlap between two texts, leading
to more judgment discrepancies. In the figure, Text B has a broader
discriminal dispersion than Text A, expanding the overlap between their
discriminal processes, a space where judges struggle to determine which
text possesses more of the trait.

Significant statistical and measurement issues can arise when the
assumption of equal dispersions between stimuli does not hold.
Specifically, the BTL model may overestimate the trait's reliability,
that is, the degree to which the outcome accurately reflects the
\emph{true} discriminal differences between stimuli. This
overestimation, in turn, results in spurious conclusions about these
differences \citep[pp.~370]{McElreath_2020} and, by extension, about the
underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
also illustrates this scenario when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the \emph{true} discriminal
difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\).

@@

Thus, erroneously assuming zero correlation between stimuli can also
lead to significant statistical and measurement issues. Specifically,
neglecting judges' biases or relevant hierarchical structures can create
dimensional mismatches in the model, leading to the over- or
underestimation of trait reliability \citep[pp.~341,
482]{Ackerman_1989, Hoyle_et_al_2023}. These inaccuracies can result in
spurious conclusions about the discriminal differences
\citep[pp.~370]{McElreath_2020} and, by extension, the underlying
discriminal processes of the stimuli. This issue is illustrated in
Figure~\ref{fig-correlation} when the discriminal difference
distribution of the BTL scores follows the thick continuous line
\((\rho = 0)\), while the \emph{true} discriminal difference follows any
discontinuous line where \(\rho \neq 0\).

Finally, as discussed in the previous section, removing \emph{misfit}
judges based risks discarding valuable information and even introduce
bias into the trait estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The direction and
magnitude of these biases remain unpredictable because they depend on
which judges researchers exclude from the analysis. \emph{Misfit judges}
are those whose evaluations deviate substantially from the shared
consensus due to the unique characteristics of either the stimuli or the
judges themselves
\citetext{\citealp[pp.~164-165]{Pollitt_2012a}; \citealp[pp.~289-290]{Pollitt_2012b}; \citealp[pp.~4]{vanDaal_et_al_2016}; \citealp[pp.~20]{Goossens_et_al_2018}}.

@@

Nevertheless, while separating the trait measurement and hypothesis
testing processes simplifies the analysis of CJ data, the statistical
literature cautions against relying solely on the point estimates of BTL
scores to conduct further analyses or hypothesis tests, as this practice
can undermine the resulting statistical conclusions. A key consideration
is that BTL scores are parameter estimates that inherently carry
uncertainty (measurement error). Ignoring this uncertainty can bias the
analysis and reduce the precision of hypothesis tests. The direction and
magnitude of such biases are often unpredictable. Results may be
attenuated, exaggerated, or remain unaffected depending on the degree of
uncertainty in the scores and the actual effects being tested
\citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
Furthermore, the reduced precision in hypothesis tests diminishes their
statistical power, increasing the likelihood of committing type-I or
type-II errors \citep{McElreath_2020}.
Figure~\ref{fig-measurement_error} illustrates these issues,
demonstrating how neglecting measurement error \((\sigma_{T})\) by
relying only on outcome averages can reduce the precision of a
predictor's estimated effect.

@@

\caption{\label{fig-measurement_error}The effect of outcome uncertainty
\(\sigma_{T}\) on the estimation of an effect \(\beta_{X}\) linked to
the predictor. The example assumes a sample size \(n=100\) and an
uncertainty that increases from left to right.}

@@

Luckily, SCMs and DAGs support identification analysis through two key
advantages. First, regardless of complexity, they can represent various
causal structures using only five fundamental building blocks. This
feature allows researchers to decompose complex structures into
manageable components, facilitating their analysis
\citep{Neal_2020, McElreath_2020}. Second, they depict causal
relationships in a non-parametric way. This flexibility enables feasible
identification strategies without requiring specification of the types
of variables, the functional forms relating them, or the parameters of
those functional forms \citep[pp.~35]{Pearl_et_al_2016}.

@@

Before incorporating the first theoretical principles of Thurstone's
theory, it is essential to further define SCMs. SCMs are formal
mathematical models characterized by a set of \emph{endogenous}
variables \(V\), a set of \emph{exogenous} variables \(E\), and a set of
functions \(F\) \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
variables are those whose causal mechanisms a researcher chooses to
model \citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other endogenous and exogenous variables.
These functions use the symbol `\(:=\)' to denote the asymmetrical
causal dependence between variables and the symbol `\(\:\bot\:\)' to
represent \emph{d-separation}, a concept akin to (conditional)
independence.

SCM~\ref{fig-cj03_scm} illustrates the inclusion of the first
theoretical principles into the conceptual-population model designed to
examine the impact of different teaching methods on students' writing
ability. This SCM outlines the relationship between the
conceptual-population outcome \((O^{cp}_{hiabjk})\) and several related
variables. The subscripts \(h\) and \(i\) label the students who
authored the texts (individuals), while the indices \(a\) and \(b\)
represent the compared texts (stimuli). The index \(j\) refers to the
judge, and \(k\) represents the judgment index, accounting for
experimental conditions where judges perform the same judgment multiple
times, i.e., a \emph{repeated measures design}
\citep[pp.~366-376]{Lawson_2015}. Notably, the indexing system
facilitates comparisons between texts written by the same student
\((h = i, a \neq b)\) and between texts written by distinct students
(\(h \neq i\), where \(a = b\) is possible). However, it excludes cases
where judges compare a student's text to itself \((h = i, a = b)\), as
such comparison lacks practical relevance within the CJ framework.

@@

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a \emph{graph}
consisting of nodes connected by edges, where nodes represent random
variables. The term \emph{directed} indicates that edges or arrows
extend from one node to another, indicating the direction of causal
influence. The absence of an edge implies no direct relationship between
the nodes. The term \emph{acyclic} means that the causal influences do
not form loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs conventionally depict observed variables as
solid black circles and unobserved (latent) variables as open circles
\citep{Morgan_et_al_2014}. They also display conditioned variables by
surrounding them with a square. Here, \emph{conditioning} refers to
restricting the statistical analysis to a subset of the population
defined by the values of the given variable \citep[pp.~32]{Neal_2020}.
Although DAGs often omit exogenous variables for simplicity, the DAGs
presented in this section includes exogenous variables to improve
clarity and reveal potential issues related to conditioning and
confounding \citep{Cinelli_et_al_2020}.
